---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# csemGT

<!-- badges: start -->
[![R-CMD-check](https://github.com/rgempp/csemGT/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rgempp/csemGT/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/rgempp/csemGT/branch/main/graph/badge.svg)](https://app.codecov.io/gh/rgempp/csemGT?branch=main)
[![CRAN status](https://www.r-pkg.org/badges/version/csemGT)](https://CRAN.R-project.org/package=csemGT)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

The goal of `csemGT` is to estimate the conditional standard error of
measurement (CSEM) within the Generalizability Theory framework for the
persons-by-items univariate single-facet crossed design, following
Brennan (1998). It was created and is maintained by
[René Gempp](https://gempp.cl) as the first of a planned series of
focused packages on CSEM, paralleling the Stata module `gtcsem`.

## Installation

You can install the development version of `csemGT` from
[GitHub](https://github.com/rgempp/csemGT) with:

``` r
# Development installation
pak::pak("rgempp/csemGT")
# or
remotes::install_github("rgempp/csemGT", build_vignettes = TRUE)
```

Once on CRAN:

``` r
install.packages("csemGT")
```

## Example

```{r example, eval = FALSE}
library(csemGT)
data(iowa_like)

fit <- csem_gt(iowa_like)
print(fit)
plot(fit, plot_type = "both", error_types = "absolute")
```

## Relationship with the forthcoming `csemR`

`csemGT` is the first of a planned family of CSEM packages. The broader
`csemR` package, currently in development, will integrate `csem_gt()` with
additional estimator families (split-half methods including Woodruff (1990),
ANOVA-based methods, and binomial/compound binomial models including Hanson's
4-parameter beta compound binomial). Users of `csemGT` will be able to migrate
to `csemR` via a simple `library()` change; the function signature and output
object are identical.

## Citation

If you use `csemGT` in published work, please cite (the entry will be
updated once the companion paper is in press):

> Gempp, R. (in preparation). *Conditional standard errors of measurement
> in generalizability theory*. Manuscript in preparation.

## References

Brennan, R. L. (1998). Raw-score conditional standard errors of measurement
in generalizability theory. *Applied Psychological Measurement*, *22*(4),
307–331. https://doi.org/10.1177/014662169802200401
