---
title: "Worked examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
library(csemGT)
```

# Overview

This vignette puts `csemGT` to work on three illustrations that
parallel, in simplified form, the worked examples in the companion
Stata paper for `gtcsem`. The polytomous extension of the first
example (the CES-D weekly scoring case in the Stata paper) is omitted
here because it depends on clinical data that are not publicly
available. In exchange, the bootstrap option is showcased below,
complementing the analytical SEs reported throughout the paper.

The three examples are:

1. A binary educational test, where we verify the equivalence of GT's
   absolute CSEM with Lord's (1955) binomial-error formula, and then
   introduce bootstrap SEs.
2. A dichotomous SAT-style test, where we compare the three relative-
   error estimators (`full`, `large_a`, `uncorrelated`) implemented
   in `csem_gt()`.
3. A Likert personality scale, where the smoother and the D-study
   features take centre stage.

The intro vignette (`vignette("intro-csem-gt", package = "csemGT")`)
covers the bare estimation workflow; readers new to the package may
prefer to start there.

# Example 1: A binary educational test, verifying Lord (1955)

In the dichotomous case, the absolute CSEM derived under
Generalizability Theory for a single-facet $p \times i$ design
collapses to Lord's (1955) classic binomial-error formula. Letting
$X_p$ denote person $p$'s observed total raw score and $n_i$ the
number of items,

$$
\hat{\sigma}(\Delta_p) \;=\; \frac{1}{n_i}\sqrt{\frac{X_p(n_i - X_p)}{n_i - 1}}.
$$

Brennan (1998) makes this equivalence explicit (his equation 8). We
verify it numerically on `iowa_like`, a simulated ITED-Vocabulary-like
dataset distributed with the package; see `?iowa_like` for the
calibration details. To keep the example fast we work with a
reproducible subsample of 600 persons.

```{r ex1-subsample}
set.seed(1998)
idx <- sample(nrow(iowa_like), size = 600)
sub <- iowa_like[idx, ]
n_i <- ncol(sub)

fit <- csem_gt(sub, error_type = "absolute", method = "full")
```

The package returns one fitted value per observed total score in
`fit$by_score`. We compare those against the Lord formula evaluated
at the same scores:

```{r ex1-lord-check}
score_tbl <- fit$by_score[, c("observed_score", "csem.absolute")]
score_tbl$csem_lord <- with(
  score_tbl,
  sqrt(observed_score * (n_i - observed_score) / (n_i - 1)) / n_i
)
score_tbl$diff <- score_tbl$csem.absolute - score_tbl$csem_lord
head(score_tbl, 8)

max(abs(score_tbl$diff))
```

The maximum absolute discrepancy is at the floor of double precision;
the two formulas are algebraically identical for the binary $p \times
i$ design.

```{r ex1-plot, fig.cap = "Per-person absolute CSEM for the iowa_like subsample (n = 600, I = 40), full estimator."}
plot(fit, main = "iowa_like: absolute CSEM by observed score")
```

## Adding bootstrap standard errors

The CSEM values just shown are per-person *point estimates*: each one
asks how precise this person's score is. A different question is how
precise *that estimate of the precision* is itself, which is a
second-order quantity. By default `csem_gt()` returns delta-method
analytical SEs and confidence intervals for it; setting `bootstrap =
TRUE` together with `return_analytical = TRUE` adds
item-resampling bootstrap counterparts in the same fit, without
displacing the analytical ones.

```{r ex1-bootstrap}
fit_boot <- csem_gt(
  sub,
  error_type        = "absolute",
  method            = "full",
  bootstrap         = TRUE,
  return_analytical = TRUE,
  R                 = 200L,
  bootstrap_type    = "item",
  ci_method         = "percentile",
  seed              = 1998
)
```

Five randomly chosen persons illustrate the side-by-side layout that
the resulting `estimates` table now carries:

```{r ex1-bootstrap-table}
cols <- c("observed_score", "csem.absolute",
          "se.analytic.absolute",     "se.boot.absolute",
          "ci_low.analytic.absolute", "ci_up.analytic.absolute",
          "ci_low.boot.absolute",     "ci_up.boot.absolute")
fit_boot$estimates[1:5, cols]
```

For the binary $p \times i$ design the two SE sources should agree to
within Monte-Carlo error; sizable discrepancies would point to
sample-size or distributional issues worth investigating. A larger
`R` will tighten the bootstrap interval; here we keep `R = 200` to
keep the vignette fast to build.

# Example 2: A dichotomous educational test, comparing estimators (SAT12)

The second example uses the SAT12 science items from package `mirt`
(Chalmers, 2012). `mirt` is a `Suggests`, so the chunks below run
only when it is installed:

```{r ex2-deps}
has_mirt <- requireNamespace("mirt", quietly = TRUE)
```

We score the multiple-choice items dichotomously against the
published key (with one corrected entry at item 32):

```{r ex2-key, eval = has_mirt}
key <- c(1, 4, 5, 2, 3, 1, 2, 1,
         3, 1, 2, 4, 2, 1, 5, 3,
         4, 4, 1, 4, 3, 3, 4, 1,
         3, 5, 1, 3, 1, 5, 4, 3)

scored <- mirt::key2binary(mirt::SAT12, key)
scored <- scored[stats::complete.cases(scored), ]
dim(scored)
```

A single `csem_gt()` call lists all three relative-error estimators at
once and disables the smoother (we want the raw scatter, not a
quadratic summary):

```{r ex2-fit, eval = has_mirt}
fit_sat <- csem_gt(
  scored,
  error_type = "relative",
  method     = c("full", "large_a", "uncorrelated"),
  smoother   = "none"
)
```

The compare layout overlays the three estimators on one panel, with
the per-person scatter (`compare_points = TRUE`) showing where they
agree, where they diverge, and where the `uncorrelated` estimator
returns negative error variance for individual respondents. Those
cases are coerced to zero by default and recorded in
`fit_sat$diagnostics`:

```{r ex2-plot, eval = has_mirt, fig.cap = "Relative CSEM under the three estimators implemented by csem_gt() for the SAT12 science test."}
plot(
  fit_sat,
  compare_methods = TRUE,
  compare_points  = TRUE,
  show_smooth     = FALSE
)
```

The `full` estimator carries within-score heterogeneity (multiple
points at the same observed score, reflecting the person-by-item
covariance term in equation 33 of Brennan, 1998); `large_a` collapses
that to a single value per score by replacing the covariance term
with its large-$A$ approximation; `uncorrelated` drops the term
entirely, which is what produces the negative-variance cases at
extreme scores.

# Example 3: A Likert personality scale, smoother and D-study

For polytomous data with non-binary response options the
`csem_gt()` interface applies without modification. We use
`ipip_like`, a simulated 10-item, 5-point Conscientiousness-like
subscale calibrated to summary statistics of the IPIP-50 inventory
(Goldberg, 1992; Goldberg, Johnson, Eber, Hogan, Ashton, Cloninger,
& Gough, 2006). See `?ipip_like` for calibration details.

```{r ex3-fit}
fit_ip <- csem_gt(
  ipip_like,
  error_type = "relative",
  method     = "full",
  smoother   = "polynomial"
)
```

The recovered variance components and the generalizability coefficient
$E\rho^2$ live in `fit_ip$variance_components`:

```{r ex3-components}
fit_ip$variance_components$person
fit_ip$variance_components$item
fit_ip$variance_components$residual
fit_ip$variance_components$reliability_coefficients$erho2
```

These match, to ANOVA tolerance, the calibration targets documented
in `?ipip_like` ($\sigma^2(p) \approx 0.434$,
$\sigma^2(i) \approx 0.136$, $\sigma^2(pi) \approx 1.000$,
$E\rho^2 \approx 0.81$). The wide per-person table can be inspected
via the data-frame method:

```{r ex3-as-df}
as.data.frame(fit_ip)[1:5, ]
```

For Likert data the within-score spread of the CSEM tends to be
larger than for binary data, and the quadratic smoother carries less
of that variation than in the educational examples. The plot with
`cibands = "model"` shows the 95% confidence band around the
quadratic fit, which is the layout used as Figure 3 of the Stata
paper:

```{r ex3-plot, fig.cap = "Relative CSEM against the observed score for ipip_like, with the 95% CI band around the quadratic fit."}
plot(fit_ip, plot_type = "both", cibands = "model")
```

## D-study: projecting to a different test length

Generalizability Theory projects the precision of a test to
alternative test lengths via the D-study (Brennan, 2001). In
`csem_gt()` this is controlled by `n_items_D`: passing it a value
different from the observed number of items rescales the error
variances and reliability coefficients to that hypothetical
length, while the G-study variance components themselves remain
fixed.

```{r ex3-dstudy}
fit_05 <- csem_gt(
  ipip_like,
  error_type = "relative", method = "full",
  smoother   = "none",
  n_items_D  = 5L
)
fit_20 <- csem_gt(
  ipip_like,
  error_type = "relative", method = "full",
  smoother   = "none",
  n_items_D  = 20L
)

rho_ref <- fit_ip$variance_components$reliability_coefficients$erho2
spearman_brown <- function(rho, K) (K * rho) / (1 + (K - 1) * rho)

dstudy <- data.frame(
  test_length = c(5L, 10L, 20L),
  erho2_gt    = c(
    fit_05$variance_components$reliability_coefficients$erho2,
    rho_ref,
    fit_20$variance_components$reliability_coefficients$erho2
  ),
  erho2_sb    = c(
    spearman_brown(rho_ref, 0.5),
    rho_ref,
    spearman_brown(rho_ref, 2.0)
  )
)
dstudy$diff <- dstudy$erho2_gt - dstudy$erho2_sb
dstudy
```

The csem_gt projection and the Spearman-Brown predicted reliability
agree closely. The small remaining differences reflect the way GT
treats the item-variance term, which Spearman-Brown's classical
derivation assumes away. For longer tests both values converge to the
same asymptote.

# References

Brennan, R. L. (1998). Raw-score conditional standard errors of
measurement in generalizability theory. *Applied Psychological
Measurement, 22*(4), 307-331.
<https://doi.org/10.1177/01466216980224002>

Brennan, R. L. (2001). *Generalizability theory*. Springer.

Chalmers, R. P. (2012). mirt: A multidimensional item response theory
package for the R environment. *Journal of Statistical Software,
48*(6), 1-29. <https://doi.org/10.18637/jss.v048.i06>

Goldberg, L. R. (1992). The development of markers for the Big-Five
factor structure. *Psychological Assessment, 4*(1), 26-42.
<https://doi.org/10.1037/1040-3590.4.1.26>

Goldberg, L. R., Johnson, J. A., Eber, H. W., Hogan, R., Ashton,
M. C., Cloninger, C. R., & Gough, H. G. (2006). The International
Personality Item Pool and the future of public-domain personality
measures. *Journal of Research in Personality, 40*(1), 84-96.
<https://doi.org/10.1016/j.jrp.2005.08.007>

Lord, F. M. (1955). Estimating test reliability. *Educational and
Psychological Measurement, 15*(4), 325-336.
<https://doi.org/10.1177/001316445501500401>

Open-Source Psychometrics Project. (n.d.). *Raw data*.
<https://openpsychometrics.org/rawdata/>
