---
title: "Introduction to csemGT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to csemGT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

## Overview

`csemGT` estimates *conditional* standard errors of measurement (CSEMs)
under Generalizability Theory for the univariate single-facet design in
which persons are crossed with items (a `p × i` design). Whereas a
single overall SEM summarises measurement precision for a whole test,
a conditional SEM describes how precision varies *along the score
scale*, so that examinees at different observed-score levels can be
assigned different error bands.

The package implements the absolute-error estimator and the three
relative-error estimators (`full`, `large_a`, `uncorrelated`) developed
by Brennan (1998), together with optional polynomial smoothing of the
conditional error variances across the score scale and analytical or
bootstrap standard errors for the per-person estimates. `csemGT` is the
R companion to the Stata `gtcsem` command and reproduces its numerical
results.

This vignette walks through a minimal analysis. For the full argument
list see `?csem_gt`; for replications of published examples see the
*Examples* vignette.

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("rgempp/csemGT")
```

The `mirt` package is an optional dependency used only by some of the
worked examples in the *Examples* vignette; it is not required for the
core functionality shown here.

## A minimal example

The package ships with `iowa_like`, a simulated `3000 × 40` matrix of
dichotomously scored item responses. It is **not** real test data: it
was generated from a Rasch model whose parameters were calibrated so
that its ANOVA-based mean error variances match the values Brennan
(1998, p. 314) reports for the ITED Vocabulary Test example. See
`?iowa_like` for the full provenance.

To keep the vignette fast we draw a reproducible subsample of 600
persons; the analysis is otherwise identical on the full matrix.

```{r fit}
library(csemGT)
data(iowa_like)

set.seed(1998)
iowa_sub <- iowa_like[sample(nrow(iowa_like), 600), ]

fit <- csem_gt(iowa_sub, error_type = "relative", method = "full")
```

`error_type = "relative"` requests the relative (δ-type) conditional
error variance, and `method = "full"` selects Brennan's finite-sample
estimator (Equations 35–36), which is the default. No bootstrap is
requested, so standard errors are the closed-form analytical
approximation.

## Inspecting the fit: `print()`

```{r print}
fit
```

The printed report is organised in blocks. The header records the
design and the estimation choices. The ANOVA table gives the
person, item, and residual mean squares and the corresponding
variance components. The D-study block reports the overall absolute
and relative error variances and SEMs: here
$\hat{\sigma}^2(\Delta) \approx .0051$ and
$\hat{\sigma}^2(\delta) \approx .0048$, closely reproducing the values
Brennan (1998) reports for the ITED Vocabulary Test and confirming that
the 600-person subsample preserves the calibrated properties of the
full dataset. The reliability-like coefficients (generalizability
coefficient $E\rho^2$ and dependability coefficient $\Phi$) and the
quadratic smoothing fit close the report.

## Per-score view: `summary()`

```{r summary}
summary(fit)
```

`summary()` adds the conditional SEM tabulated by observed score,
together with each score group's size, cumulative frequency, and
percentile. Reading down the `rel_full` column shows the
characteristic concave pattern: the relative CSEM is small at the
extremes of the score scale, rises to a maximum near the middle, and
declines again — the same shape Brennan (1998, Figure 1d) obtained for
the fitted relative SEM of a dichotomously scored test.

## Visualising: `plot()`

```{r plot}
plot(fit)
```

The default plot shows the per-person relative conditional SEM against
observed score, overlaid with the fitted quadratic smooth. Additional
layouts — comparing estimators in a single panel, side-by-side panels,
and analytical or bootstrap confidence bands — are documented in
`?plot.csem` and demonstrated in the *Examples* vignette.

## The fitted object

`csem_gt()` returns an object of class `"csem"`. Its main components
are the per-person estimates (`$estimates`), the by-score table
(`$by_score`), the G-study variance components
(`$variance_components`), the smoothing fits (`$smooth_fits`), and the
smoother sample diagnostics (`$diagnostics`). Convenience accessors are
provided:

```{r accessors}
str(fit, max.level = 1)

# variance components
coef(fit)

# by-score table as a data frame
head(by_score(fit))
```

## Next steps

- `?csem_gt` documents every argument, including bootstrap options,
  polynomial smoothing, extreme-score handling, and the `n_items_D`
  D-study generalisation.
- The *Examples* vignette replicates published analyses, including a
  multi-panel comparison of the relative-error estimators.
- The Stata `gtcsem` command produces the same estimates for users
  working in that environment.

## References

Brennan, R. L. (1998). Raw-score conditional standard errors of
measurement in generalizability theory. *Applied Psychological
Measurement, 22*(4), 307–331.
